Xolotl: An Intuitive and Approachable Neuron \& Network Simulator in MATLAB

### 2018-07-27 paper/paper.tex
Add citations and begin editing benchmarks section

Two major approaches have dominated the design of neuron simulators. First, some simulators, such as "NEURON" (\cite{hinesNEURONSimulationEnvironment1997}) are written in a fast compiled language like "C" and allow for the construction and simulation of neuron models using object-oriented paradigms. These simulators tend to perform fast computations with little overhead, but suffer from a steep learning curve. Wrapping these simulators in a more approachable language like "Python" and the use of graphical user interfaces (GUIs) mitigate these drawbacks, at the cost of obfuscating the underlying algorithms and parameters (\cite{bretteSimulationNetworksSpiking2007, hinesNEURONPython2009}). Another approach is to design simulators that can be used entirely in popular scientific programming languages. These simulators can be easier to use, and can benefit from inter-compatibility with other tools. Programs like "DynaSim" (\cite{sherfeyDynaSimMATLABToolbox2018}), "ANNarchy" (\cite{vitayANNarchyCodeGeneration2015}), "BRIAN" (\cite{stimbergBrianSecondComing2013}) allow the user to specify models using strings of equations or using components that are specified in the scientific programming language, that can then be translated into a faster implementation language such as "C" or "C++" (\cite{stimbergEquationorientedSpecificationNeural2014}). This approach permits considerable flexibility for simulating systems of differential equations. Since models need to be translated between the two languages, the hierarchical nature of neuron models is not naturally encapsulated by these tools, and the syntax can be verbose. Neither approach facilitates the creation of tools that simultaneously maintain efficiency, ease-of-use, and clarity.

@@

Here, \( \bar{g}_{j} \) is the maximal conductance, and \( E_{j} \) is the reversal potential of the ion channel population \( j \). \( A_{i} \) is the area of compartment \( i\) that contains these ion channels.  \( m_{j} \) and \( h_{j} \) are activation and inactivation variables that change according to

@@

Typically, \( \tau_{m} \), \( \tau_{h} \), \( m_{\infty} \), and \( h_{\infty} \) are functions of the membrane potential \( V_{i} \). The software uses integration schemes that have been specifically developed to solve equations of this form (cite D\&A, Hines, others), though other schemes can be used if desired.

@@

\paragraph{New mechanisms require new \texttt{C++} code.} Adding new network components requires writing new "C++" code. Adding a new conductance in the Hodgkin-Huxley formalism (\cite{hodgkinMeasurementCurrentvoltageRelations1952, dayanTheoreticalNeuroscience2001}), requires creating a new "C++" header file, though this is generally trivial. Implementing a new integration scheme or component type requires much more in-depth knowledge of the underlying "C++" core code.

@@

The axon of the giant squid contains contains a fast inactivating sodium conductance ("NaV"), a slower non-inactivating potassium conductance ("Kd"), and a passive leak current. Seminal work by Hodgkin and Huxley showed that depolarizations of the membrane, for example, by injecting current, could lead to an activation of the voltage-sensitive "NaV" channels, which led to a run away depolarization that was terminated by the inactivation of "NaV" channels and the activation of "Kd" channels that repolarized the membrane (\cite{hodgkinComponentsMembraneConductance1952, hodgkinMeasurementCurrentvoltageRelations1952}). As one of the simplest models of excitable neural mebranes, the Hodgkin-Huxley model often serves as a the first model introduced in pedagogical literature (cite some textbooks).

@@

The "manipulate" function opens the GUI which permits visualization of changing parameters in real-time. Moving sliders representing the values of network parameters updates a plot (Figure \ref{fig:puppeteerscreenshot}B). By default, the function opens a figure displaying the results of the "plot" function, which shows the voltage and intracellular calcium traces for each compartment (Figure \ref{fig:puppeteerscreenshot}A). "manipulate" grants slider control over all "xolotl" parameters by default, but specific ones can be selected by passing them as arguments.

@@

\begin{figure}
	\centering
	\includegraphics[width=1.0\linewidth]{gfx/figure_benchmark}
	\caption{\texttt{xolotl} benchmarked against \texttt{DynaSim} and \texttt{NEURON}. (A) Ratio of 'simulated' time to runtime (speed factor) and accuracy, measured by spike train coincidence plotted against decreasing time-resolution. (B) Speed factor for models at increasing simulation times. (C) Speed factor over number of compartments.}
	\label{fig:figurebenchmark}
\end{figure}

>>>>

Two major approaches have dominated the design of neuron simulators. First, some simulators, such as "NEURON" (\cite{hinesNEURONSimulationEnvironment1997}) are written in a fast compiled language like "C" and allow for the construction and simulation of neuron models using object-oriented paradigms. These simulators tend to perform fast computations with little overhead, but suffer from a steep learning curve. Wrapping these simulators in a more approachable language like "Python" and the use of graphical user interfaces (GUIs) mitigate these drawbacks (\cite{hinesNEURONPython2009}), at the cost of obfuscating the underlying algorithms and parameters (\cite{bretteSimulationNetworksSpiking2007, hinesNEURONPython2009}). Another approach is to design simulators that can be used entirely in popular scientific programming languages. These simulators can be easier to use, and can benefit from inter-compatibility with other tools. Programs like "DynaSim" (\cite{sherfeyDynaSimMATLABToolbox2018}), "ANNarchy" (\cite{vitayANNarchyCodeGeneration2015}), "BRIAN" (\cite{stimbergBrianSecondComing2013}), "morphforge" (\cite{hullMorphforgeToolboxSimulating2014}), "PyNN" (\cite{davisonPyNNCommonInterface2009}) allow the user to specify models using strings of equations or using components that are specified in the scientific programming language, that can then be translated into a faster implementation language such as "C" or "C++" (\cite{stimbergEquationorientedSpecificationNeural2014}). This approach permits considerable flexibility for simulating systems of differential equations. Since models need to be translated between the two languages, the hierarchical nature of neuron models is not naturally encapsulated by these tools, and the syntax can be verbose. Neither approach facilitates the creation of tools that simultaneously maintain efficiency, ease-of-use, and clarity.

@@

Here, \( \bar{g}_{j} \) is the maximal conductance, and \( E_{j} \) is the reversal potential of the ion channel population \( j \). \( A_{i} \) is the surface area of compartment \( i\) that contains these ion channels.  \( m_{j} \) and \( h_{j} \) are activation and inactivation variables that change according to

@@

Typically, \( \tau_{m} \), \( \tau_{h} \), \( m_{\infty} \), and \( h_{\infty} \) are functions of the membrane potential \( V_{i} \). The software uses integration schemes that have been specifically developed to solve equations of this form (\cite{dayanTheoreticalNeuroscience2001, hinesEfficientComputationBranched1984, ohErrorAnalysisSpecialized2006}), though other schemes can be used if desired.

@@

\paragraph{New mechanisms require new \texttt{C++} code.} Adding new network components requires writing new "C++" code. Adding a new conductance in the Hodgkin-Huxley formalism (\cite{hodgkinComponentsMembraneConductance1952, hodgkinMeasurementCurrentvoltageRelations1952, hodgkinQuantitativeDescriptionMembrane1952, dayanTheoreticalNeuroscience2001}), requires creating a new "C++" header file, though this is generally trivial. Implementing a new integration scheme or component type requires much more in-depth knowledge of the underlying "C++" core code.

@@

The axon of the giant squid contains contains a fast inactivating sodium conductance ("NaV"), a slower non-inactivating potassium conductance ("Kd"), and a passive leak current. Seminal work by Hodgkin and Huxley showed that depolarizations of the membrane, for example, by injecting current, could lead to an activation of the voltage-sensitive "NaV" channels, which led to a run away depolarization that was terminated by the inactivation of "NaV" channels and the activation of "Kd" channels that repolarized the membrane (\cite{hodgkinComponentsMembraneConductance1952, hodgkinMeasurementCurrentvoltageRelations1952}). As one of the simplest models of excitable neural mebranes, the Hodgkin-Huxley model often serves as a the first model introduced in pedagogical literature (\cite{dayanTheoreticalNeuroscience2001, sterrattPrinciplesComputationalModelling2011, trappenbergFundamentalsComputationalNeuroscience2010}).

@@

The "manipulate" function opens the GUI which permits visualization of changing parameters in real-time. Moving sliders representing the values of network parameters updates a plot (Figure \ref{fig:figuremanipulate}B). By default, the function opens a figure displaying the results of the "plot" function, which shows the voltage and intracellular calcium traces for each compartment (Figure \ref{fig:figuremanipulate}A). "manipulate" grants slider control over all "xolotl" parameters by default, but specific ones can be selected by passing them as arguments.

@@

\begin{figure}
	\centering
	\includegraphics[width=1.0\linewidth]{gfx/figure_benchmark}
	\caption{\texttt{xolotl} benchmarked against \texttt{DynaSim} and \texttt{NEURON}. The top row benchmarks using a tonically-firing Hodgkin-Huxley model with 3 conductances with constant injected current. The bottom row repeats the same benchmarks using a bursting stomatogastric ganglion model with 8 conductances. (A,E) Ratio of run-time to simulation time (speed factor) over increasing integration time step. (B,F) Error over increasing integration time step as computed by the LeMasson matrix where the baseline condition is the voltage trace at $\Delta t = 0.001$ ms. (C,G) Simulations over increasing simulation time where $\Delta t = 0.1$ ms. (D,H) Simulations over increasing number of unconnected compartments.}
	\label{fig:figurebenchmark}
\end{figure}
